\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-01-12 honors_thesis/journal_paper/paper.tex
I think I am finally done with revisions to my honors thesis/GECCO paper

Because the motion of particles in PSO depends only on the \emph{positions} of
the particle and its neighbors, not the \emph{value} of the objective function
at those positions, speculative evaluation is possible.  We do not have to wait
for a particle's position to be evaluated to know where it might move to next,
so we can compute the set of all possible next positions and send them to be
evaluated along with the particle's current position.  Once the current
position of each particle has been evaluated, we can determine which of the
speculatively evaluated positions matches the branch taken by the particle.  We
have thus evaluated two iterations at once.  This can easily get unwieldy if
the set of next possible positions for each particle is large, but a wise
choice of topology limits that set to a reasonable size.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Table~\ref{tab:evals} lists all possible combinations of updates
that can be seen by a particle with two neighbors other than itself, which is
the case in all of the topologies we use.  If all of these speculative
positions are evaluated in parallel along with the original, non-speculative
position of the particle at time $t-1$, then one of the speculative positions
has to be the correct one, and two iterations have been evaluated at once.

\begin{table}
  \caption{All possible updates for a particle with two neighbors}
  \label{tab:evals}
  \centering
  \begin{tabular}{c|c}
	$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	No update&No update\\
	Updated&No update\\
	No update&Neighbor 1\\
	No update&Neighbor 2\\
	Updated&The particle itself\\
	Updated&Neigbhor 1\\
	Updated&Neighbor 2\\
  \end{tabular}
\end{table}

Because $2(n+1)$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require on
the order of $p^2$ speculative evaluations per iteration, where $p$ is the
swarm size.  Clearly it is much more desirable to have a sparse topology, where
$O(np)$ is much smaller than $O(p^2)$.  However, some functions are better
optimized with the Complete topology and the quick spread of information it
entails than with sparse topologies.  Accordingly, we use the Random topology
described in~\cite{mcnabb-cec09}, which has been shown to effectively simulate
the Complete topology.  In that topology, 2 neighbors are chosen at random from
the swarm for each particle at each iteration.  Because a particle is also
always its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report
the results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

@@

We ran experiments to compare our speculative PSO algorithm to the standard
PSO algorithm.  To use the same number of processors for each algorithm,
speculative evaluation needs a smaller swarm size than standard PSO.  For each
of the topologies we used, a particle has two neighbors other than itself.  As
shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
the number of particles in the speculative swarm.  The extra evaluations done
in speculative evaluation would instead be particles in regular PSO, so we
compare swarms of size $p_s$ in speculative evaluation with swarms of size
$8p_s$ in regular PSO.

We use a Random topology in our speculative algorithm, as the Complete topology
leads to an explosion of speculative evaluations.  If speculative evaluation
were not being performed, it is possible that the Complete topology would be
used.  However, the Complete topology also requires a very large amount of
interprocessor communication in parallel PSO, so it is still quite possible
that Random would be used even with standard PSO.  But, to be fair in our
comparisons, we compare to the standard PSO algorithm using both the Random
topology with two neighbors and the Complete topology.  All functions have 20
dimensions, and all results shown are averages over 20 runs.  We use 240
processors in our experiments, so our speculative algorithm uses 30 particles
and standard PSO uses 240 particles.  For ease of labeling, we call our
speculative algorithm SEPSO and standard PSO just PSO, with the topology
following.

@@

In performing speculative evaluation as we have described it, $2(n+1)$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  We can try to make use of those evaluations instead of
throwing them away.

@@

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2(n+1)$.  While we cannot with certainty predict the correct branch every
time, we can look at the statistics of the behavior of particles in general and
perhaps come up with a good guess, reducing the number of necessary speculative
evaluations.

@@

The statistics for the Complete topology are less promising (as there are 240
neighbors that branch four splits into, instead of two), though we do not use
that topology with speculative evaluation anyway.

Because there are a few branches with very high probabilities, we can have hope
that cutting out some of the evaluations that have low probability will lead to
an increase in overall performance.

@@

@@

>>>>

Speculative evaluation is possible in PSO because the motion of particles
depends only on the \emph{positions} of the particle and its best neighbor, not
the \emph{value} of the objective function at those positions.  We do not have
to wait for a particle's function evaluation to complete in order to know where
it might move to next, so we can compute the set of all possible next positions
and send them to be evaluated along with the particle's current position.  Once
the current position of each particle has been evaluated, we can determine
which of the speculatively evaluated positions matches the branch taken by the
particle.  We have thus evaluated two iterations at once.  This can easily get
unwieldy if the set of next possible positions for each particle is large, but
a wise choice of topology limits that set to a reasonable size.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
all possible combinations of updates that can be seen by a particle with two
neighbors other than itself, which is the case in all of the topologies we use.
If all of these speculative positions are evaluated in parallel along with the
original, non-speculative position of the particle at time $t-1$, then one of
the speculative positions has to be the correct one, and two iterations have
been evaluated at once.

\begin{table}
  \caption{All possible updates for a particle with two neighbors}
  \label{tab:evals}
  \centering
  \begin{tabular}{cc}
	$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	\hline
	No update&No update\\
	\hline
	No update&Neighbor 1\\
	\hline
	No update&Neighbor 2\\
	\hline
	Updated&No update\\
	\hline
	Updated&Neighbor 1\\
	\hline
	Updated&Neighbor 2\\
	\hline
	Updated&The particle itself\\
	\hline
  \end{tabular}
\end{table}

Because $2(n+1)-1$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require
$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In that topology, 2 neighbors are chosen at random from the
swarm for each particle at each iteration.  Because a particle is also always
its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

@@

We ran experiments to compare our speculative PSO algorithm to the standard
PSO algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  The speculative algorithm actually performs two iterations of
PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
evaluations.  For benchmark functions with very fast evaluations this may not
be the best use of parallelism in PSO.  But for many real-world applications,
the objective function takes on the order of minutes or more to evaluate; in
such cases our assumption is perfectly reasonable.

To use the same number of processors for each algorithm, speculative evaluation
needs a smaller swarm size than standard PSO.  For each of the topologies we
used, a particle has two neighbors other than itself.  As shown in
Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
the number of particles in the speculative swarm.  The extra evaluations done
in speculative evaluation would instead be particles in regular PSO, so we
compare swarms of size $p_s$ in speculative evaluation with swarms of size
$8p_s$ in regular PSO.

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion of speculative
evaluations.  If speculative evaluation were not being performed, it is
possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
parallel PSO, so it is still quite possible that Random would be used even with
standard PSO.  But, to be fair in our comparisons, we compare to the standard
PSO algorithm using both the Random topology with two neighbors and the
Complete topology.  All functions have 20 dimensions, and all results shown are
averages over 20 runs.  We use 240 processors in our experiments, so our
speculative algorithm uses 30 particles and standard PSO uses 240 particles.
For ease of labeling, we call our speculative algorithm SEPSO and standard PSO
just PSO, with the topology following.

@@

In performing speculative evaluation as we have described it, $2(n+1)-1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  We can try to make use of those evaluations instead of
throwing them away.

@@

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2(n+1)-1$.  While we cannot with certainty predict the correct branch every
time, we can look at the statistics of the behavior of particles in general and
perhaps come up with a good guess, reducing the number of necessary speculative
evaluations.

@@

The statistics for the Complete topology are less promising, as there are 240
neighbors that branch four splits into, instead of two.  Pruning does, however,
allow for the possibility of using the Complete topology in speculative
evaluation while avoiding the explosion in the number of evaluations it would
otherwise entail.  Though pruning makes its use possible, we still found the
communication costs in our parallel environment prohibitive, so we do not
present any results for the Complete topology here.

With the other topologies, however, there are a few branches with very high
probabilities, so we can have hope that cutting out some of the evaluations
that have low probability will lead to an increase in overall performance.

@@

We need not simply produce speculative children for the next iteration.  We can
view all possible speculative evaluations for a particle as an infinite tree
with branching factor $2(n+1)-1$.  As we have already seen that doing one full
level of the tree produces too many extra evaluations to be profitable, it is
incredibly unlikely that doing two full iterations would produce decent
results.  But, if the idea of speculating more than one iteration ahead is
combined with wisely pruning the possible evaluations based on branch
statistics, we can use just a few extra evaluations to go two or three
iterations ahead on the most likely branches.

The question of which branches to take in this infinite tree is an intriguing
one that we can only begin to explore here.  If the branch corresponding to
stagnation has a 75\% chance of being taken, as in the Random topology with
most functions, we could speculate three iterations ahead on that branch and
still have a 42\% chance of predicting correctly.  However, intuition would say
that perhaps it is better to hope that the particle is productive instead of
stagnant, so a branch where the particle updates its personal best might be
more fruitful to try.  In our experiments we try just one of the countless
possibilities, but that one turns out to work extremely well.  More work is
needed to compare the different branching possibilities on various functions.

@@

One issue remains.  What if we really want a swarm of 50 particles, but we have
230 processors to use?  The methods we have thus far presented do the same
thing for every particle, and thus would require some multiple of 50
processors---we would have 30 extra processors with nothing to do.  It might
very well be the best idea to simply adjust the swarm size slightly, but we
present another method that allows particles to behave differently, some doing
more evaluations than others.

If we use the concept of tokens commonly used in multirobot stuff (see, for
example, \dots), we can specify a number of tokens available for particles to
use, where each token corresponds to a speculative evaluation.  Whichever
method is used to which evaluations to do, they are prioritized so that
particles with only one or two tokens perform the most valuable speculative
evaluations.  Particles that perform well keep their tokens, particles that
perform poorly send one token (if they have one) to another random particle in
the swarm.

We choose to define ``perform poorly'' as a failure to update the personal best
when the original particle's position is evaluated.  Clearly there are other
possible methods of deciding when to give up tokens, but many of the methods
that sound the best are difficult in distributed environments, like giving
tokens to that particles that most recently found the globally best positions.
The use of a central machine to decide which particles get tokens would help to
facilitate such methods.

There are thus two distinct uses for tokens.  First, they could be used to try
to evenly spread extra evaluations due to an incapatible number of particles
and processors.  In cursory experimentation, we found that when only going one
iteration ahead keeping the number of evaluations per particle as even as
possible worked much better than trying to prioritize particles and let some
guess correctly more often because they have more tokens. (horrible explanation\dots)

The other, more compelling reason to use tokens is when speculating many
iterations ahead.  Particles that are going in good directions, as evidenced by
relatively frequent updates to the particle's personal best, keep their tokens
and are able to speculate up to three or four iterations.  Particles that
aren't doing as well lose their tokens and probably contribute to the
exploration of the algorithm, for which speculative evaluation doesn't help
much.

As in the previous section, we can only offer a very preliminary exploration of
the possibilities of using tokens to improve the allocation of speculative
evaluations to particles.
